---
title: "Alpha diversity"
author: "YOUR NAME HERE"
date: "`r format(Sys.time())`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

This markdown is to calculate the alpha diversity of the samples as richness (number of species detected per sample) 
and as Shannon diversity index (accounts for number of species detected and the evenness of their abundance). 
These calculations are based on the species abundance table. Here we will plot richness and Shannon index,
 and we will compare these diversity metrics between the study groups.


### Packages
Loading libraries needed. They were installed in the codespace when developed.
```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(patchwork)
library(vegan)
library(FSA)
```

# Load data objects

## Load species abundance, taxonomical annotation and metadata file
```{r}
# Abundance table
abundance <- readRDS(file = "data/MetaphlanAbundance_Species.rds")
rownames(abundance) <- gsub("_SRR_db1.metaphlan", "", rownames(abundance))
dim(abundance)

# Taxonomical annotation
annotation <- readRDS(file = "data/MetaphlanAnnotations_Species.rds")
dim(annotation)

# Metadata
metadata <- read.table(file = "data/metadata.tsv", header = TRUE,
                       sep = "\t",
                       quote = "",
                       row.names = NULL)
rownames(metadata) <- metadata$Sample
dim(metadata)

# Take only samples with sequencing and metadata information
abundance <- abundance[rownames(metadata), ]
dim(abundance)
```

## Creating a directory for the PCoA related results
```{r}
results_dir <- "results/report/01_Alpha-diversity"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Calculate richness and shannon
```{r}
# We calculate richness and Shannon index using vegan package 
# and added both to a dataframe
alpha_div <- data.frame(
  Richness = rowSums(abundance > 0),
  Shannon = vegan::diversity(abundance, index = "shannon"),
  row.names = rownames(abundance)
)
```

# Plot alpha diversity measurements

## Prepare plot dataframe
```{r}
# We join the alpha diversity data with the grouping file
# (often called metadata)
plot_data <- cbind(alpha_div, metadata$Area, metadata$Sampling,
  paste(metadata$Area, metadata$Sampling, sep = "_")
)

colnames(plot_data) <-
  c("Richness", "Shannon", "Area", "Sampling_Date", "Area_Sampling")

write.csv(plot_data, file = "results/report/01_Alpha-diversity/01_Alpha_diversity.csv",
          row.names = TRUE, quote = FALSE)
```

## Create boxplots 
```{r}

colors <- c("1_M2" = "#349F2C", "1_M3" = "#BAE4B3",
            "2_M2" = "#0F5096", "2_M3" = "#1779E3",
            "3_M2" = "#CA0020", "3_M3" = "#FF8784")

<<<<<<< Updated upstream
richness_boxplot <- ggplot(plot_data, 
                           aes(x = Sampling_Date, y = Richness)) +
=======
richness_boxplot <- ggplot(plot_data, aes(x = Sampling_Date, y = Richness)) +
>>>>>>> Stashed changes
  geom_boxplot(aes(fill = Area_Sampling), outliers = FALSE, alpha = 0.7) +
  geom_point(aes(fill = Area_Sampling), shape = 21,
             position = position_jitter(width = 0.25, height = 0)) +
  theme_bw() +
  theme(legend.position = "none", panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_manual(values = colors) +
  labs(x = "Sampling", y = "Richness (species detected)") +
  facet_grid(. ~ Area)

<<<<<<< Updated upstream


```


## Shannon


```{r}

=======
>>>>>>> Stashed changes
shannon_boxplot <- ggplot(plot_data, aes(x = Sampling_Date, y = Shannon)) +
  geom_boxplot(aes(fill = Area_Sampling), outliers = FALSE, alpha = 0.7) +
  geom_point(aes(fill = Area_Sampling), shape = 21,
             position = position_jitter(width = 0.25, height = 0)) +
  theme_bw() +
  theme(legend.position = "none", panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_fill_manual(values = colors) +
  labs(x = "Sampling", y = "Shannon index") +
  facet_grid(. ~ Area)

richness_boxplot + shannon_boxplot
<<<<<<< Updated upstream
```

### Save

```{r}
=======

>>>>>>> Stashed changes
ggsave(filename = "results/report/01_Alpha-diversity/02_alphadiv_boxplots.png",
       width = 8, height = 4, dpi = 300)
```

# Statistical comparison
<<<<<<< Updated upstream

We have three areas that we want to compare,
we will use a non-parametric test for more than 2 groups: Kruskal-Wallis.
We are testing whether there are differences among the three areas

```{r}
kruskal.test(Richness ~ Area, plot_data)
```

** Is Richness  significantly different among the areas? **

When a Kruskal-Wallis test shows that there are significant differences 
between groups, you need a post-hoc test to determine which specific groups
differ. Since Kruskal-Wallis is non-parametric, the appropriate post-hoc 
test is also anon-parametric - dunnTest. We correct for multiple testing 
using Benjamini-Hochberg (BH) procedure, which is a method used to control 
the False Discovery Rate (FDR) in multiple hypothesis testing

# Perform post-hoc Dunn's test

```{r}
dunnTest(Richness ~ Area, plot_data, method = "bh")
```

**Differences occur between 1-3 and 2-3**

Is it the same for Shannon?
```{r}
kruskal.test(Shannon ~ Area, plot_data)
```

**Almost significant!**

When it is not significant, we don't needed to test between groups, but lets double check:

```{r}
dunnTest(Shannon ~ Area, plot_data, method = "bh")
```

**No significant differences, but comparison of 1-3 is almost significant.**



=======
```{r}
# We have three areas that we want to compare,
# we will use a non-parametric test for more than 2 groups: Kruskal-Wallis.
# We are testing whether there are differences among the three areas

kruskal.test(Richness ~ Area, plot_data)
# Richness is significantly different among the areas!

# When a Kruskal-Wallis test shows that there are significant differences 
# between groups, you need a post-hoc test to determine which specific groups
# differ. Since Kruskal-Wallis is non-parametric, the appropriate post-hoc 
# test is also anon-parametric - dunnTest. We correct for multiple testing 
# using Benjamini-Hochberg (BH) procedure, which is a method used to control 
# the False Discovery Rate (FDR) in multiple hypothesis testing

# Perform post-hoc Dunn's test
dunnTest(Richness ~ Area, plot_data, method = "bh")
# Differences occur between 1-3 and 2-3

# Same for Shannon
kruskal.test(Shannon ~ Area, plot_data)
# Almost significant!

# Not significant, so not needed to test between groups, but lets double check:
dunnTest(Shannon ~ Area, plot_data, method = "bh")
# No significant differences, but comparison of 1-3 is almost significant.
```

>>>>>>> Stashed changes
# Printing all package versions (good practice to ensure reproducibility)
```{r session-info, echo=FALSE}
# Print R version and loaded packages
sessionInfo()
```
