---
title: "Abundance overview"
author: "YOUR NAME HERE"
date: "`r format(Sys.time())`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

This markdown is to plot an abundance overview of the top-10 most abundant species of the dataset

# R Packages that need to be installed
```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(RColorBrewer)
```

# Load data objects

## Load species abundance, taxonomical annotation and metadata file
```{r}
# Abundance table
abundance <- readRDS(file = "data/MetaphlanAbundance_Species.rds")
rownames(abundance) <- gsub("_SRR_db1.metaphlan", "", rownames(abundance))
dim(abundance)

# Taxonomical annotation
annotation <- readRDS(file = "data/MetaphlanAnnotations_Species.rds")
dim(annotation)

# Metadata
metadata <- read.table(file = "data/metadata.tsv",
                       header = TRUE, sep = "\t",
                       quote = "",
                       row.names = NULL)
rownames(metadata) <- metadata$Sample
dim(metadata)

# Take only samples with sequencing and metadata information
abundance <- abundance[rownames(abundance) %in% rownames(metadata),]
dim(abundance)
```

# Creating a directory for the PCoA related results
```{r}
results_dir <- "results/report/03_Abundance-overview"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Calculate total abundances per species
```{r}
taxa_total <- colSums(abundance)
top_taxa <- names(sort(taxa_total, decreasing = TRUE))[1:10]
```

# Filter to top 10 taxa
```{r}
abundance_top10 <- as.data.frame(abundance) %>% select(all_of(top_taxa))
```

# Convert to long format for ggplot
```{r}
# Move rownames (samples) to a column
abundance_top10$Sample <- rownames(abundance_top10)

abundance_long <- abundance_top10 %>%
  pivot_longer(cols = -Sample, names_to = "Species",
               values_to = "Abundance")

# Adding the species name for the plot
abundance_long$Species_name <- annotation$species[match(abundance_long$Species,
                                                  rownames(annotation))]

# Adding the grouping columns: Area and Sampling
abundance_long$Area <- metadata$Area[match(abundance_long$Sample,
                                           metadata$Sample)]
abundance_long$Sampling <- metadata$Sampling[match(abundance_long$Sample,
                                               metadata$Sample)]
```

# Stack bar plot of the top-10 most dominant species
```{r}
# Ensure Sample is a factor ordered by Sampling + Sample name
# (for visual grouping)
abundance_long <- abundance_long %>%
  arrange(Sampling, Sample) %>%
  mutate(Sample = factor(Sample, levels = unique(Sample)))

pbar <- ggplot(abundance_long, aes(x = Sample, y = Abundance,
                                   fill = Species_name)) +
  geom_bar(stat = "identity", position = "stack", show.legend = TRUE) +
  scale_fill_brewer(palette = "Paired") +
  labs(title = "Top-10 Most Abundant Species",
       x = "Sample",
       y = "Relative Abundance (%)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0)) +
  facet_wrap(. ~ Area, scales = "free")

pbar

# Save it as a png
ggsave("results/report/03_Abundance-overview/01_top10_species.png",
       width = 8, height = 6, dpi = 300)
```

# Let's summarize the same plot per Area and Sampling
```{r}
# Add Sample column and merge with metadata
abundance_top10$Area <- metadata$Area
abundance_top10$Sampling <- metadata$Sampling

# Pivot to long format
abundance_long <- abundance_top10 %>%
  pivot_longer(cols = all_of(top_taxa),
               names_to = "Species",
               values_to = "Abundance")

# Summarize abundance by Area + Sampling + Species
abundance_summary <- abundance_long %>%
  group_by(Area, Sampling, Species) %>%
  summarise(MeanAbundance = mean(Abundance), .groups = "drop")

# Adding the species name for the plot
abundance_summary$Species_name <- 
  annotation$species[match(abundance_summary$Species,
                           rownames(annotation))]

pbar_sum <- ggplot(abundance_summary, 
                   aes(x = interaction(Area, Sampling, sep = " | "),
                       y = MeanAbundance, fill = Species_name)) +
  geom_bar(stat = "identity", position = "stack", show.legend = TRUE) +
  scale_fill_brewer(palette = "Paired") +
  theme_bw() +
  labs(title = "Mean Abundance of Top 10 Taxa by Area and Sampling",
       x = "Area | Sampling",
       y = "Mean Relative Abundance (%)") +
  theme(axis.text.x = element_text(hjust = 0.5),
        strip.text = element_text(face = "bold")) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0))

pbar_sum

# Save it as a png
ggsave("results/report/03_Abundance-overview/02_top10_species_per_area_sampling.png",
       width = 8, height = 6, dpi = 300)

```

# Printing all package versions (good practice to ensure reproducibility)
```{r session-info, echo=FALSE}
# Print R version and loaded packages
sessionInfo()
```