---
title: "Alpha diversity"
author: "YOUR NAME HERE"
date: "`r format(Sys.time())`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
params:
  markdown_id: "Alpha diversity"
  data: "data/species_table.tsv"
  groups: "data/Groups.tsv"
  data_title: "Relative species abundance"
  do_paired_test: FALSE
  id_variable: NULL
editor_options: 
  chunk_output_type: console
---

This markdown is to calculate the alpha diversity of the samples as richness (number of species detected per sample) and as Shannon diversity index (accounts for number of species detected and the evenness of their abundance). These calculations are based on the species abundance. Here we will plot richness and Shannon index and we will compare between the study groups.


### Packages
```{r packages, message=FALSE, warning=FALSE}

#library(cowplot)
#library(dplyr)
library(ggplot2)
#library(glue)
#library(magrittr)
library(patchwork)
#library(picante)
#library(scales)
#library(stringr)
#library(tibble)
#library(usethis)
library(vegan)
library(FSA)
#library(tidylog, warn.conflicts = FALSE) # tidylog should be loaded after dplyr and tidyr and conflicts should be silenced since it (purposely) builds on top of those packages
```

# Load data objects

## Load species abundance and grouping file
```{r}
# Abundance table
abundance <- readRDS(file="data/MetaphlanAbundance_Species.rds")
rownames(abundance) <- gsub("_SRR_db1.metaphlan", "", rownames(abundance))
dim(abundance)

# Taxonomical annotation
annotation <- readRDS(file="data/MetaphlanAnnotations_Species.rds")
dim(annotation)

# Metadata
metadata <- read.table(file="data/metadata.tsv", header = T, sep = '\t', quote = "", row.names=NULL)
rownames(metadata) <- metadata$Sample
dim(metadata)

# Take only samples with sequencing and metadata information
abundance <- abundance[rownames(abundance) %in% rownames(metadata),]
dim(abundance)
```

## Creating a directory for the PCoA related results
```{r}
results_dir <- "results/report/01_Alpha-diversity"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

```

# Calculate richness and shannon
```{r}
# We calculate richness and Shannon index using vegan package and added both to a dataframe
alphaDiv <- data.frame(
  Richness = rowSums(abundance > 0),
  Shannon = vegan::diversity(abundance, index = "shannon"),
  row.names = rownames(abundance))
```

# Plot alpha diversity measurements
## Prepare plot dataframe
```{r}
# We join the alpha diversity data with the grouping file (often called metadata)
plot_data <- cbind(alphaDiv, metadata$Area, metadata$Sampling, 
paste(metadata$Area, metadata$Sampling, sep="_"))
str(plot_data)
head(plot_data)
colnames(plot_data) <- c("Richness", "Shannon", "Area", "Sampling_Date", "Area_Sampling")

write.table(plot_data, file = "results/report/01_Alpha-diversity/01_Alpha_diversity.xlsx", row.names = T, col.names = T, sep = "\t", quote = F)
```

## Create boxplots 
```{r, fig.width= 8, fig.height= 4}

colors = c("1_M2" = "#349F2C", "1_M3" = "#BAE4B3",  
  "2_M2" = "#0F5096", "2_M3" = "#1779E3", 
  "3_M2" = "#CA0020", "3_M3" = "#FF8784")

richness_boxplot <- ggplot(plot_data, aes(x = Sampling_Date, y = Richness)) +
  geom_boxplot(aes(fill = Area_Sampling), outliers = F, alpha = 0.7) +
  geom_point(aes(fill = Area_Sampling), shape = 21, 
             position = position_jitter(width = 0.25, height = 0)) +
  theme_bw() +
  theme(legend.position = "none", panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()) +
  scale_fill_manual(values = colors) +
  labs(x = "Sampling", y = "Richness (species detected)") +
  facet_grid(.~Area)

shannon_boxplot <- ggplot(plot_data, aes(x = Sampling_Date, y = Shannon)) +
  geom_boxplot(aes(fill = Area_Sampling), outliers = F, alpha = 0.7) +
  geom_point(aes(fill = Area_Sampling), shape = 21, 
             position = position_jitter(width = 0.25, height = 0)) +
  theme_bw() +
  theme(legend.position = "none", panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  scale_fill_manual(values = colors) +
  labs(x = "Sampling", y = "Shannon index") +
  facet_grid(.~Area)

richness_boxplot + shannon_boxplot

ggsave(filename = "results/report/01_Alpha-diversity/02_alphadiv_boxplots.pdf", width = 8, height = 4)
```

# Statistical comparison
```{r}
# We have three areas that we want to compare, we will use a non-parametric test for more than 2 groups: Kruskal-Wallis.
# We are testing whether there are differences between groups
kruskal.test(Richness ~ Area, plot_data)

# When a Kruskal-Wallis test shows that there are significant differences between groups, 
# you need a post-hoc test to determine which specific groups differ. 
#Since Kruskal-Wallis is non-parametric, the appropriate post-hoc test is also non-parametric.

# Perform post-hoc Dunn's test
dunnTest(Richness ~ Area, plot_data, method = "bh")

# Same for Shannon 
kruskal.test(Shannon ~ Area, plot_data)
# Not significant, so not needed but lets double check:
dunnTest(Shannon ~ Area, plot_data, method = "bh")
```

# Printing all package versions (good practice to ensure reproducibility)
```{r session-info, echo=FALSE}
# Print R version and loaded packages
sessionInfo()
```
