---
title: "Beta diversity - PCoA"
author: "YOUR NAME HERE"
date: "`r format(Sys.time())`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

This markdown is to calculate Bray-Curtis dissimilarities between samples based on species abundance and to plot these ones on the two first axis of a principal coordinate analysis (PCoA).

# R Packages that need to be installed
```{r packages, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(vegan)
```

# Load data objects

## Load species abundance, taxonomical annotation and metadata file
```{r}
# Abundance table
abundance <- readRDS(file = "data/MetaphlanAbundance_Species.rds")
rownames(abundance) <- gsub("_SRR_db1.metaphlan", "", rownames(abundance))
dim(abundance)

# Taxonomical annotation
annotation <- readRDS(file = "data/MetaphlanAnnotations_Species.rds")
dim(annotation)

# Metadata
metadata <- read.table(file = "data/metadata.tsv",
                       header = TRUE, sep = "\t",
                       quote = "",
                       row.names = NULL)
rownames(metadata) <- metadata$Sample
dim(metadata)

# Take only samples with sequencing and metadata information
abundance <- abundance[rownames(metadata), ]
dim(abundance)
```

# Creating a directory for the PCoA related results
```{r}
results_dir <- "results/report/02_Beta-diversity"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Plotting a Principal coordinates analysis (PCoA)

## Step 1: Calculate Bray-Curtis Dissimilarity Matrix
```{r}
bray_curtis <- vegdist(abundance, method = "bray")
```

## Step 2: Perform PCoA
```{r}
pcoa_result <- cmdscale(bray_curtis, eig = TRUE, k = 2)
```


## Extract eigenvalues and points
```{r}
eig_vals <- pcoa_result$eig
# Only positive eigenvalues contribute to variance explained
eig_vals <- eig_vals[eig_vals > 0]
var_explained <- eig_vals / sum(eig_vals)
axis_variance <- round(var_explained[1:2] * 100)

scores <- as.data.frame(pcoa_result$points)
scores$Area <- metadata$Area
scores$Sampling <- metadata$Sampling
colnames(scores) <- c("Axis1", "Axis2", "Area", "Sampling")
```

## Step 3: Extract loading data
```{r}
# To calculate variable loadings (contributions of the original variables 
# to the PCoA axes), you can use the envfit() function in the vegan package.
# Fit environmental vectors to the ordination
loadings <- envfit(pcoa_result, abundance, perm = 999)

loadings_df <- as.data.frame(loadings$vectors$arrows[loadings$vectors$r >= 0.75 & 
            loadings$vectors$pvals <= 0.05,])
colnames(loadings_df) <- c("Axis1", "Axis2")
```

# Step 4: Plot PCoA results
```{r}
colors <- c("1" = "#349F2C",
            "2" = "#0F5096",
            "3" = "#CA0020")
scores$Area <- as.factor(scores$Area) # as a factor

shapes <- c("M2" = 21,
            "M3" = 17)
scores$Sampling <- as.factor(scores$Sampling) # as a factor

p <- ggplot(scores, aes(x = Axis1, y = Axis2)) +
  geom_point(aes(color = Area, fill = Area, shape = Sampling), size = 3) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  labs(x = paste0("Axis1 ", axis_variance[1], "%"),
       y = paste0("Axis2 ", axis_variance[2], "%"),
       title = "PCoA plot")
p

# Save it as a png
ggsave("results/report/02_Beta-diversity/01_PCoA_plot.png",
       width = 8, height = 6, dpi = 300)
```

## Step 5: Plot the PCoA with Loadings (optional)
```{r}
p <- ggplot(scores, aes(x = Axis1, y = Axis2)) +
  geom_point(aes(color = Area, fill = Area, shape = Sampling), size = 2) +
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0, xend = Axis1, yend = Axis2),
               arrow = arrow(length = unit(0.1, "cm")), color = "grey",
               linewidth = 0.5) +
  geom_text(data = loadings_df, aes(x = Axis1, y = Axis2,
            label = annotation$species[match(rownames(loadings_df),
            rownames(annotation))]),
            vjust = -0.5,
            hjust = 0.5,
            color = "red",
            position = "jitter",
            size = 3) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme_bw() +
  theme(aspect.ratio = 1) +
  labs(x = paste0("Axis1 ", axis_variance[1], "%"),
       y = paste0("Axis2 ", axis_variance[2], "%"),
       title = "PCoA with variable loadings",
       subtitle = "r >= 0.75, P <= 0.05")
p

# Save it as a png
ggsave("results/report/02_Beta-diversity/02_PCoA_with_loadings.png",
       width = 8, height = 6, dpi = 300)
```

# Permutational multivariate analysis of variance (PERMANOVA)
Before running a PERMANOVA test let's assess the data dispersion within the areas
and within sampling dates. High dispersion of data within the groups could affect 
the analysis of variance test as this test assumes homogeneity of variance 
## Dispersion for Area

### Distance to centroid (dispersion)
```{r}

dispersion <- betadisper(d = bray_curtis,
                         group = metadata$Area,
                         type = "centroid",
                         bias.adjust = FALSE)
dispersion_table <- permutest(dispersion,
                              permutations = how(nperm = 1000),
                              by = "margin")

dispersion_table
# There is a significant dispersion of the data points among the areas
```

### Dispersion boxplots
```{r}
dispersion_plot_data <- data.frame("distance" = dispersion$distances,
                                   "group" = dispersion$group)

dispersion_box <- ggplot(dispersion_plot_data, aes(x = group, y = distance)) +
  geom_boxplot(aes(fill = group), outliers = FALSE, alpha = 0.7) +
  geom_point(aes(fill = group), shape = 21,
             position = position_jitter(width = 0.25, height = 0)) +
  theme_bw() +
  scale_fill_manual(values = colors) +
  labs(title = "Dispersion (Area variability)",
       x = "Area",
       y = "Distance to centroid")

dispersion_box

# Save it as a png
ggsave("results/report/02_Beta-diversity/03_dispersion_plot.png",
       width = 6, height = 4, dpi = 300)
```

## Dispersion for Sampling

### Distance to centroid (dispersion)
```{r}

dispersion <- betadisper(d = bray_curtis,
                         group = metadata$Sampling,
                         type = "centroid",
                         bias.adjust = FALSE)
dispersion_table <- permutest(dispersion,
                              permutations = how(nperm = 1000),
                              by = "margin")

dispersion_table
# There is a significant dispersion of the data points between the sampling dates
```

### Dispersion boxplots
```{r}
dispersion_plot_data <- data.frame("distance" = dispersion$distances,
                                   "group" = dispersion$group)

dispersion_box <- ggplot(dispersion_plot_data, aes(x = group, y = distance)) +
  geom_boxplot(aes(fill = group), outliers = FALSE, alpha = 0.7) +
  geom_point(aes(fill = group), shape = 21,
             position = position_jitter(width = 0.25, height = 0)) +
  theme_bw() +
  scale_fill_manual(values = c("red", "blue")) +
  labs(title = "Dispersion (Sampling variability)",
       x = "Sampling",
       y = "Distance to centroid")

dispersion_box

# Save it as a png
ggsave("results/report/02_Beta-diversity/04_dispersion_plot_sampling.png",
       width = 5, height = 4, dpi = 300)
```

# PERMANOVA test
A significant permutest from a betadisper object indicates that there are 
significant differences in dispersion among groups, which violates the 
assumption of homogeneity of variances required for PERMANOVA. 
You can still perform a PERMANOVA, but the results should be interpreted 
with caution, and you should consider additional analyses.
```{r}
# Differences among treatment groups
set.seed(12)

# # TODO: complain if not at least color group OR shape group is given
# With all data
independent_variables <- paste("Area", "Sampling", sep = "+")

form <- paste0("bray_curtis ~", independent_variables) %>% as.formula
form

adonis_table <- adonis2(form,
                        data = metadata,
                        permutations = 1000,
                        by = "margin")

adonis_table
# Both factors: area and sampling data significantly explain the compositional 
# variance among the samples. These results should be interpreted with caution 
# as there is significant differences in dispersion among the groups.

# Let's save the statistical summary table:
write.csv(adonis_table,
          file = "results/report/02_Beta-diversity/05_Adonis_table.csv",
          row.names = TRUE, quote = FALSE)
```

# Printing all package versions (good practice to ensure reproducibility)
```{r session-info, echo=FALSE}
# Print R version and loaded packages
sessionInfo()
```

