---
title: "Beta diversity - PCoA"
author: "YOUR NAME HERE"
date: "`r format(Sys.time())`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
params:
  markdown_id: "Beta-diversity-PCoA"
  show_x_labels: TRUE
  dist_method: "bray" 
  data: "../data/species_table.tsv"
  groups: "../data/Groups.tsv"
  data_title: "Relative species abundance"
editor_options: 
  chunk_output_type: console
---

This markdown is to calculate Bray-Curtis dissimilarities between samples based on species abundance and to plot these ones on the two first axis of a principal coordinate analysis (PCoA).

# R Packages that need to be installed
```{r packages, message=FALSE, warning=FALSE}
library(dplyr)
library(tidylog, warn.conflicts = FALSE) # tidylog should be loaded after dplyr and tidyr and conflicts should be silenced since it (purposely) builds on top of those packages
library(ggplot2)
library(vegan)
library(readxl)
library(plotly)
```

# Load data objects

## Load species abundance and grouping file
```{r}
# Abundance table
data <- read.table(params$data, header = T, sep = '\t', quote = "")
dim(data)

# Grouping table
groups <- read.table(params$groups, header = T, sep = '\t', quote = "")
dim(groups)

# Take only samples with sequencing and metadata information
data <- data[rownames(data) %in% groups$Samples,]
dim(data)
```

# Creating a directory for the PCoA related results
```{r}
results_dir <- "../results/report/Beta-diversity"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

```

# Building the Principal coordinates analysis (PCoA)
```{r, results='asis'}
##### PCoA with loadings
## Step 1: Calculate Bray-Curtis Dissimilarity Matrix
# Compute Bray-Curtis dissimilarity matrix
bray_curtis <- vegdist(data, method = params$dist_method)

## Step 2: Perform PCoA
# Perform PCoA
pcoa_result <- cmdscale(bray_curtis, eig = TRUE, k = 2)

# Extract eigenvalues and points
#pcoa_result2 <- cmdscale(bray_curtis, eig = TRUE, k = nrow(data) - 1)
eig_vals <- pcoa_result$eig
# Only positive eigenvalues contribute to variance explained
eig_vals <- eig_vals[eig_vals > 0]
var_explained <- eig_vals / sum(eig_vals)
axis_variance <- round(var_explained[1:2] * 100)

scores <- as.data.frame(pcoa_result$points)
scores$Group <- groups$Groups
colnames(scores) <- c("Axis1", "Axis2", "Group")

## Step 3: Perform PCoA
#To calculate variable loadings (contributions of the original variables to the PCoA axes), you can use the envfit() function in the vegan package.
# Fit environmental vectors to the ordination
loadings <- envfit(pcoa_result, data, perm = 999)

# Extract loading data
loadings_df <- as.data.frame(loadings$vectors$arrows[loadings$vectors$r >= 0.50 & loadings$vectors$pvals <= 0.05,])
colnames(loadings_df) <- c("Axis1", "Axis2")

## Step 4: Plot the PCoA with Loadings
# Plot PCoA results
colors = c("#989994", "#1779E3")

p <- ggplot(scores, aes(x = Axis1, y = Axis2)) +
  geom_point(aes(color = Group), size = 2) +
  geom_segment(data = loadings_df, 
               aes(x = 0, y = 0, xend = Axis1, yend = Axis2),
               arrow = arrow(length = unit(0.1, "cm")), color = "grey", 
               size = 0.5) +
  geom_text(data = loadings_df, 
            aes(x = Axis1, y = Axis2, label = rownames(loadings_df)),
            vjust = -0.5, 
            hjust = 0.5, 
            color = "red",
            position = "jitter",
            size = 2) +
  scale_color_manual(values = colors) +
  theme(aspect.ratio = 1) +
  theme_minimal() +
  labs(x = paste0("Axis1 ", axis_variance[1], "%"), 
       y = paste0("Axis2 ", axis_variance[2], "%"), 
       title = "PCoA with variable loadings", 
       subtitle = "r >= 0.50, P <= 0.05") # this is hardcoded for now, not ideal!
  
p

# Save it as a pdf
ggsave("../results/report/Beta-diversity/PCoA_plot.pdf", width = 8, height = 6)

# Let's make it an interactive plot with the ggplotly library
plotly_p <- ggplotly(p)
plotly_p

# Save as an interactive HTML file
library(htmlwidgets)
saveWidget(as_widget(plotly_p), "../results/report/Beta-diversity/PCoA_plot.html")
```

# Permutational multivariate analysis of variance (PERMANOVA)
## Distance to centroid (dispersion)

```{r}

dispersion <- betadisper(d = bray_curtis, group = groups$Groups, type = "centroid", bias.adjust = F)
dispersion_table <- permutest(dispersion, permutations = how(nperm = 1000), by = "margin")

dispersion_table
```

## Dispersion boxplots

```{r}

dispersion_plot_data <- data.frame("distance" = dispersion$distances,
                                   "group" = dispersion$group)

dispersion_box <- ggplot(dispersion_plot_data, aes(x = group, y = distance)) +
  geom_boxplot(aes(fill = group), outliers = F, alpha = 0.7) +
  geom_point(aes(fill = group), shape = 21, 
             position = position_jitter(width = 0.25, height = 0)) +
  theme_minimal() +
  scale_fill_manual(values = colors) +
  labs(title = "Dispersion (Group variability)", x = "Group", y = "Distance to centroid")
  
dispersion_box

ggsave("../results/report/Beta-diversity/dispersion_plot.pdf", width = 5, height = 4)
```

# PERMANOVA test (cannot run for phylogenetic distance metrics)

```{r}
# Differences among treatment groups
set.seed(12)

# # TODO: complain if not at least color group OR shape group is given
# With all data
independent_variables <- c("Groups")
 
form <- paste0("bray_curtis ~", independent_variables) %>% as.formula
form

adonis_table <- adonis2(form, 
                        data = groups,
                        permutations = 1000,
                        by = "margin")

adonis_table

```

# Printing all package versions (good practice to ensure reproducibility)
```{r session-info, echo=FALSE}
# Print R version and loaded packages
sessionInfo()
```

