---
title: "Beta diversity - PCoA"
author: "YOUR NAME HERE"
date: "`r format(Sys.time())`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

This markdown is to calculate Bray-Curtis dissimilarities between samples based on species abundance and to plot these ones on the two first axis of a principal coordinate analysis (PCoA).

# R Packages that need to be installed
```{r packages, message=FALSE, warning=FALSE}
library(dplyr)
library(tidylog, warn.conflicts = FALSE) # tidylog should be loaded after dplyr and tidyr and conflicts should be silenced since it (purposely) builds on top of those packages
library(ggplot2)
library(vegan)
library(readxl)
library(plotly)
```

# Load data objects

## Load species abundance, taxonomical annotation and metadata file
```{r}
# Abundance table
abundance <- readRDS(file = "data/MetaphlanAbundance_Species.rds")
rownames(abundance) <- gsub("_SRR_db1.metaphlan", "", rownames(abundance))
dim(abundance)

# Taxonomical annotation
annotation <- readRDS(file = "data/MetaphlanAnnotations_Species.rds")
dim(annotation)

# Metadata
metadata <- read.table(file = "data/metadata.tsv",
                       header = TRUE, sep = "\t",
                       quote = "",
                       row.names = NULL)
rownames(metadata) <- metadata$Sample
dim(metadata)

# Take only samples with sequencing and metadata information
abundance <- abundance[rownames(abundance) %in% rownames(metadata),]
dim(abundance)
```

# Creating a directory for the PCoA related results
```{r}
results_dir <- "results/report/02_Beta-diversity"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Plotting a Principal coordinates analysis (PCoA)
```{r, fig.width = 8, fig.height = 6}
##### PCoA with loadings
## Step 1: Calculate Bray-Curtis Dissimilarity Matrix
# Compute Bray-Curtis dissimilarity matrix
bray_curtis <- vegdist(abundance, method = "bray")

## Step 2: Perform PCoA
# Perform PCoA
pcoa_result <- cmdscale(bray_curtis, eig = TRUE, k = 2)

# Extract eigenvalues and points
eig_vals <- pcoa_result$eig
# Only positive eigenvalues contribute to variance explained
eig_vals <- eig_vals[eig_vals > 0]
var_explained <- eig_vals / sum(eig_vals)
axis_variance <- round(var_explained[1:2] * 100)

scores <- as.data.frame(pcoa_result$points)
scores$Area <- metadata$Area
scores$Sampling <- metadata$Sampling
colnames(scores) <- c("Axis1", "Axis2", "Area", "Sampling")

## Step 3: Extract loading data
# To calculate variable loadings (contributions of the original variables to the PCoA axes), 
# you can use the envfit() function in the vegan package.
# Fit environmental vectors to the ordination
loadings <- envfit(pcoa_result, abundance, perm = 999)

loadings_df <- as.data.frame(loadings$vectors$arrows[loadings$vectors$r >= 0.75 & 
            loadings$vectors$pvals <= 0.05,])
colnames(loadings_df) <- c("Axis1", "Axis2")

# Step 4: Plot PCoA results
colors <- c("1" = "#349F2C",
            "2" = "#0F5096",
            "3" = "#CA0020")
scores$Area <- as.factor(scores$Area) # as a factor

shapes <- c("M2" = 21,
            "M3" = 17)
scores$Sampling <- as.factor(scores$Sampling) # as a factor

p <- ggplot(scores, aes(x = Axis1, y = Axis2)) +
  geom_point(aes(color = Area, fill = Area, shape = Sampling), size = 3) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme(aspect.ratio = 1) +
  theme_minimal() +
  labs(x = paste0("Axis1 ", axis_variance[1], "%"),
       y = paste0("Axis2 ", axis_variance[2], "%"),
       title = "PCoA plot")
p

# Save it as a pdf
ggsave("results/report/02_Beta-diversity/01_PCoA_plot.pdf",
       width = 8, height = 6)

## Step 5: Plot the PCoA with Loadings (optional)
p <- ggplot(scores, aes(x = Axis1, y = Axis2)) +
  geom_point(aes(color = Area, fill = Area, shape = Sampling), size = 2) +
  geom_segment(data = loadings_df,
               aes(x = 0, y = 0, xend = Axis1, yend = Axis2),
               arrow = arrow(length = unit(0.1, "cm")), color = "grey",
               size = 0.5) +
  geom_text(data = loadings_df, aes(x = Axis1, y = Axis2,
            label = annotation$species[match(rownames(loadings_df),
            rownames(annotation))]),
            vjust = -0.5,
            hjust = 0.5,
            color = "red",
            position = "jitter",
            size = 3) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  scale_shape_manual(values = shapes) +
  theme(aspect.ratio = 1) +
  theme_minimal() +
  labs(x = paste0("Axis1 ", axis_variance[1], "%"),
       y = paste0("Axis2 ", axis_variance[2], "%"),
       title = "PCoA with variable loadings",
       subtitle = "r >= 0.75, P <= 0.05")
p

# Save it as a pdf
ggsave("results/report/02_Beta-diversity/02_PCoA_with_loadings.pdf",
       width = 8, height = 6)

# Let's make it an interactive plot with the ggplotly library (optional)
plotly_p <- ggplotly(p)
plotly_p

# Convert to JSON and save
json_str <- plotly_json(plotly_p, jsonedit = FALSE)
cat(json_str,
    file = "results/report/02_Beta-diversity/03_PCoA_plot_interactive.json")
```

# Permutational multivariate analysis of variance (PERMANOVA)

## Distance to centroid (dispersion)
```{r}

dispersion <- betadisper(d = bray_curtis,
                         group = groups$Groups,
                         type = "centroid",
                         bias.adjust = FALSE)
dispersion_table <- permutest(dispersion,
                              permutations = how(nperm = 1000),
                              by = "margin")

dispersion_table
```

## Dispersion boxplots

```{r}

dispersion_plot_data <- data.frame("distance" = dispersion$distances,
                                   "group" = dispersion$group)

dispersion_box <- ggplot(dispersion_plot_data, aes(x = group, y = distance)) +
  geom_boxplot(aes(fill = group), outliers = FALSE, alpha = 0.7) +
  geom_point(aes(fill = group), shape = 21,
             position = position_jitter(width = 0.25, height = 0)) +
  theme_minimal() +
  scale_fill_manual(values = colors) +
  labs(title = "Dispersion (Group variability)",
       x = "Group",
       y = "Distance to centroid")
  
dispersion_box

ggsave("../results/report/Beta-diversity/dispersion_plot.pdf", width = 5, height = 4)
```

# PERMANOVA test (cannot run for phylogenetic distance metrics)

```{r}
# Differences among treatment groups
set.seed(12)

# # TODO: complain if not at least color group OR shape group is given
# With all data
independent_variables <- c("Groups")
 
form <- paste0("bray_curtis ~", independent_variables) %>% as.formula
form

adonis_table <- adonis2(form, 
                        data = groups,
                        permutations = 1000,
                        by = "margin")

adonis_table

```

# Printing all package versions (good practice to ensure reproducibility)
```{r session-info, echo=FALSE}
# Print R version and loaded packages
sessionInfo()
```

