---
title: "Abundance overview"
author: "YOUR NAME HERE"
date: "`r format(Sys.time())`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

This markdown is to plot an abundance overview of the top-10 most abundant species of the dataset

# R Packages that need to be installed
```{r packages, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(RColorBrewer)
library(FSA)
```

# Load data objects

## Load species abundance, taxonomical annotation and metadata file
```{r}
# Abundance table
abundance <- readRDS(file = "data/MetaphlanAbundance_Species.rds")
rownames(abundance) <- gsub("_SRR_db1.metaphlan", "", rownames(abundance))
dim(abundance)

# Taxonomical annotation
annotation <- readRDS(file = "data/MetaphlanAnnotations_Species.rds")
dim(annotation)

# Metadata
metadata <- read.table(file = "data/metadata.tsv",
                       header = TRUE, sep = "\t",
                       quote = "",
                       row.names = NULL)
rownames(metadata) <- metadata$Sample
dim(metadata)

# Take only samples with sequencing and metadata information
abundance <- abundance[rownames(metadata),]
dim(abundance)
```

# Creating a directory for the PCoA related results
```{r}
results_dir <- "results/report/04_Group-comparison"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

# Areas comparison
As there are three areas, we will use a non-parametric test (not assuming any distribution)
for three groups: Kruskal-Wallis test.
```{r}
# We need to run this test for all species comparing among Areas (1, 2, 3)
# We need therefore a loop to run all the test and later we need to correct 
# for multiple testing (BH). FOr the significant ones we will run a pot-hoc 
# dunnTest to figure out between which groups the differences exist.

# Merge abundance data with metadata
abundance_df <- as.data.frame(cbind(abundance,
                                    Area = metadata$Area,
                                    Sample = metadata$Sample))

# Get taxa columns
taxa_cols <- setdiff(colnames(abundance_df), c("Sample", "Area"))
length(taxa_cols) # 294 species - 294 hypothesis testing!

# Let's apply a prevalence filter as my abundance table is very sparse 
# and I want to reduce the noise of species only found in a sampl.
# Let's keep only taxa present (non-zero abundance) in at least 20% of samples.
# Step 1: Prevalence Filter — keep taxa present in ≥20% of samples
n_samples <- nrow(abundance_df)
min_prevalence <- 0.20 * n_samples

# Calculate prevalence per taxon
# Subset only the taxa part of the data (numeric matrix)
abundance_taxa_only <- abundance_df[, taxa_cols]

prevalent_taxa <- sapply(abundance_taxa_only[, taxa_cols],
                         function(x) sum(x > 0)) >= min_prevalence
filtered_taxa <- names(prevalent_taxa[prevalent_taxa])
length(filtered_taxa) # 130 species passed the filter - 
#130 hypothesis for testing!

# Step 1: Kruskal–Wallis test for each taxon
# A more elegant way to run loops in R is using one of its "apply" functions
kruskal_results <- lapply(filtered_taxa, function(taxon) {
  test_data <- abundance_df[, c(taxon, "Area")]
  colnames(test_data) <- c("Abundance", "Area")
  kruskal <- kruskal.test(Abundance ~ Area, data = test_data)
  data.frame(
    Taxon = taxon,
    p_value = kruskal$p.value,
    statistic = kruskal$statistic
  )
})

# Combine results
kruskal_df <- do.call(rbind, kruskal_results)
rownames(kruskal_df) <- kruskal_df$Taxon

# Adjusting for multiple testing by Bonferroni and Hochberg method
# (False Discovery rate)
kruskal_df$adj_p_value <- p.adjust(kruskal_df$p_value, method = "BH")

# Filter significant taxa (e.g., FDR < 0.05)
significant_taxa <- kruskal_df %>% filter(adj_p_value < 0.05)
significant_taxa
dim(significant_taxa)
significant_taxa$Species_name <-
  annotation$species[match(rownames(significant_taxa),
                           rownames(annotation))]

# Step 2: Run Dunn’s test for significant taxa
dunn_results <- lapply(significant_taxa$Taxon, function(taxon) {
  test_data <- abundance_df[, c(taxon, "Area")]
  colnames(test_data) <- c("Abundance", "Area")
  test_data$Abundance <- as.numeric(test_data$Abundance)

  # Run dunnTest only if we have >1 group with non-zero data
  tryCatch({
    dunn <- dunnTest(Abundance ~ Area, data = test_data, method = "bh")
    dunn_df <- dunn$res
    dunn_df$Taxon <- taxon
    return(dunn_df)
  }, error = function(e) {
    message(paste("Skipping", taxon, "due to error:", e$message))
  })
})

# Combine Dunn test results
dunn_df <- do.call(rbind, dunn_results)
dunn_df$Species_name <-
  annotation$species[match(dunn_df$Taxon,
                           rownames(annotation))]

# Let's save the statistics for Kruskal-Wallis test 
# ordered by adj_p_value and with more logical order of the columns
significant_taxa <- significant_taxa %>%
  arrange(adj_p_value) %>%
  relocate(Taxon, Species_name, statistic, p_value, adj_p_value)

write.table(significant_taxa, 
    file = "results/report/04_Group-comparison/01_kruskal-wallis-results.xlsx",
    row.names = TRUE, col.names = TRUE, sep = "\t", quote = FALSE)

# Let's save the post-hoc Dunn test results
write.table(dunn_df,
  file = "results/report/04_Group-comparison/02_post-hoc-dunn-results.xlsx",
  row.names = TRUE, col.names = TRUE, sep = "\t", quote = FALSE)
```

# Plotting the top-20 significant species
```{r}
# 1. Get top 20 taxa from Kruskal–Wallis
top_taxa <- kruskal_df %>%
  arrange(adj_p_value) %>%
  slice(1:20) %>%
  pull(Taxon)

# 2. Prepare long-format data with top taxa
abundance_df$Sampling <- metadata$Sampling
abundance_long <- abundance_df %>%
  select(Sample, Area, Sampling, all_of(top_taxa)) %>%
  pivot_longer(
    cols = all_of(top_taxa),
    names_to = "Taxon",
    values_to = "Abundance"
    ) 

# 3. Order taxa by significance and fix data-types
abundance_long$Abundance <- as.numeric(abundance_long$Abundance)
abundance_long$Species_name <-
  annotation$species[match(abundance_long$Taxon,
                           rownames(annotation))]
###
abundance_long$Taxon <- factor(abundance_long$Taxon, levels = top_taxa)

# 4. Plot heatmap with ggplot2
heatmap_top20 <- ggplot(abundance_long, 
                        aes(x = Sample, y = Taxon,
                        fill = log10(Abundance+0.000001))) +
  geom_tile(color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "C", name = "Abundance (log10)") +
  facet_grid(. ~ Area, scales = "free") +
  theme_bw() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 10),
    axis.title = element_blank()
  )

heatmap_top20

# Save it as a pdf
ggsave("results/report/04_Group-comparison/03_significants_heatmap.pdf",
       width = 8, height = 8)
```

# Printing all package versions (good practice to ensure reproducibility)
```{r session-info, echo=FALSE}
# Print R version and loaded packages
sessionInfo()
```