---
title: "Microbial Association Networks using NetCoMi"
author: "Sebastian Ayala Ruano"
date: "`r format(Sys.time())`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: show
    df_print: paged
editor_options: 
  chunk_output_type: console
---

This markdown is to create microbial association networks (MANs) based on the abundance data using the the NetCoMi package.

## Work environment setup 

### Import R Packages
```{r packages, message=FALSE, warning=FALSE}
library(NetCoMi)
```

### Load data
```{r}
# Abundance table
abundance <- readRDS(file = "data/MetaphlanAbundance_Species.rds")
rownames(abundance) <- gsub("_SRR_db1.metaphlan", "", rownames(abundance))
dim(abundance)

# Taxonomical annotation
annotation <- readRDS(file = "data/MetaphlanAnnotations_Species.rds")
dim(annotation)

# Metadata
metadata <- read.table(file = "data/metadata.tsv",
                       header = TRUE, sep = "\t",
                       quote = "",
                       row.names = NULL)
rownames(metadata) <- metadata$Sample
dim(metadata)

# Take only samples with sequencing and metadata information
abundance <- abundance[rownames(abundance) %in% rownames(metadata),]
dim(abundance)

# Replace the column names of the abundance table with the species names
colnames(abundance) <- annotation$species[match(colnames(abundance), rownames(annotation))]
```

### Create a directory to store the results
```{r}
results_dir <- "results/report/04_Microbial-association-networks"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)
```

## Microbial association network (MAN) with CCLasso as association measure

We create a MAN with the [Correlation inference for Compositional data through Lasso (CCLasso)](https://netcomi.de/reference/cclasso) 
method. **CCLasso** estimates positive definite Pearson correlations from log-ratio variances using 
a **latent variable model** based on least squares with an L1 penalty. 
[Aitchison](https://doi.org/10.1111/j.2517-6161.1982.tb01195.x) introduced log-ratios as a basis for analyzing 
compositional data, as these values are independent of the total sum of counts. Consequently, CCLasso addresses 
the compositional nature of metagenomics data.

### Network construction

The [`netConstruct()`](https://netcomi.de/reference/netconstruct) function is used to build the network. 
To make the calculations feasible, we filter the taxa to the 50 most abundant species using the `filtTax` 
and `filtTaxPar` parameters.

Normalization is performed internally as part of the CCLasso method, so we set `normMethod` to `none`. 
To handle zeros in the data, we applied the `pseudoZO` method, which adds a pseudo count of 1 to zero values.

The `signed` method was aplied for transforming associationss into dissimilarities, so that 
**strongly negatively associated taxa** have a high dissimilarity and, in turn, a low similarity, 
which corresponds to **edge weights** in the network plot.

A threshold of `0.5` is applied as **sparsification method**, meaning that only species with a correlation 
greater than or equal to 0.5 will be included in the network.

The **verbose** argument is set to `3` so that all messages generated by `netConstruct()` as well as messages 
of external functions are printed. Finally, we set a **seed** for reproducibility.

```{r}
net_cclasso <- netConstruct(abundance,
                           measure = "cclasso",
                           filtTax = "highestFreq",
                           filtTaxPar = list(highestFreq = 50),
                           normMethod = "none",
                           zeroMethod = "pseudoZO",
                           dissFunc = "signed",
                           sparsMethod = "threshold",
                           thresh = 0.5,
                           verbose = 3,
                           seed = 2236)
```

Now, we can check and save the **edge list** of the network. It contains the estimated associations, dissmiliarities, and 
similarities (`adj` column) for each pair of species. 

```{r}
head(net_cclasso$edgelist1)

write.table(net_cclasso$edgelist1, 
            file = file.path(results_dir, "1_cclasso_network_edgelist.csv"),
            sep = ",", quote = FALSE, row.names = FALSE)
```

We can also use the`plotHeat` function to plot the estimated associations.

```{r}
# Open PNG device
png(file.path(results_dir, "2_cclasso_network_association_heatmap.png"), width = 2000, height = 1800, res = 150)

# Generate the plot
plotHeat(net_cclasso$assoMat1, textUpp = "none", textLow = "none")

# Close device
dev.off()
```


### Network analysis  

The [`netAnalyze()`](https://netcomi.de/reference/netanalyze) function is used for analyzing the MAN.

Here, **centrLCC** is set to `FALSE`, so the centralities are calculated for all nodes in the network, 
not only for those in the **largest connected component (LCC)**. **weightDeg** and **normDeg** are set to 
`FALSE`, so that the degree of a node is simply defined as the number of connections it has to other nodes.

**Hubs** are defined here as nodes with a degree above the 90% quantile of the empirical degree distribution,
which is specified by the **hubPar** and **hubQuant** arguments.

Clusters are identified using the **greedy modularity optimization** algorithm 
([`cluster_fast_greedy`](https://r.igraph.org/reference/cluster_fast_greedy.html)). 
This algorithm optimize the modularity of networks by iteratively grouping nodes into communities, 
aiming to maximize the density of intra-community edges compared to inter-community edges

```{r}
props_cclasso <- netAnalyze(net_cclasso, 
                            # Centrality
                            centrLCC = FALSE,
                            weightDeg = FALSE,
                            normDeg = FALSE,
                            
                            # Hubs
                            hubPar = "degree",
                            hubQuant = 0.9,
                            
                            # Clustering
                            clustMethod = "cluster_fast_greedy"
                    )

# Print the summary of the network analysis
summary(props_cclasso, numbNodes = 5L)
```

### Network visualization

Finally, we can visualize the network using the [`plot()`](https://netcomi.de/reference/plot) function.

```{r}
# Open PNG device
png(file.path(results_dir, "3_cclasso_network.png"), width = 2500, height = 2300, res = 150)

# Generate the plot
plot(props_cclasso,
     # Network
     layout = "spring",
     repulsion = 0.9,
     title1 = "CCLasso Network",
     showTitle = TRUE,
     rmSingles = TRUE,
     cexTitle = 2.3,

     # Nodes 
     nodeColor = "cluster", 
     nodeSize = "degree",
     cexNodes = 2.7, 
     nodeSizeSpread = 2,
     shortenLabels = "simple",
     labelLength = 8L,

     # Edges
     posCol = "#ac1214", # red
     negCol ="#0d1fad", # blue
)

legend(0.7, 1.1, title = "Estimated associations:", legend = c("+","-"), 
        col = c("#ac1214","#0d1fad"), inset = 0.02, cex = 1.7, lty = 1, lwd = 3.5, 
        bty = "n", horiz = TRUE, y.intersp = 0.7)

# Close device
dev.off()
```